---
title: "Lecture 05: Map Practice and Synthesis"
author: "Hadley Stabnau"
format: pdf
editor: visual
---

## Problem 1

# New packages
 `rnaturalearth`
 `rnaturalearthdata`
 `sf`
 `mapview`
 
 
#  New functions 
 
  * plot() will give us a plot with the data and then differentiate plots between the arguments inside. 
 ex. `plot(data(type = "countries"))`
 this will differentiate the plots between the different types of countries.  
 
 * A better function for these graphs is mapview::mapview()
 
 ## Problem 2
 
 # Why does the plot made with the following code say "0 countries"?
```{r}
ggplot(data = world) +
    geom_sf() +
    xlab("Longitude") + ylab("Latitude") +
    ggtitle("World map", subtitle = paste0("(", length(unique(world$NAME)), " countries)"))
```
 
 * This is because of the capital in Name on line 4.
 * the fixed code looks like
 
```{r}
ggplot(data = world) +
    geom_sf() +
    xlab("Longitude") + ylab("Latitude") +
    ggtitle("World map", subtitle = paste0("(", length(unique(world$name)), " countries)"))
```
 
 
## Problem 3

```{r}
ggplot(data = world) + 
  geom_sf(fill= "antiquewhite") + 
  geom_text(data= world_points,
            aes(x=X, y=Y, label=name), 
            color = "darkblue", 
            fontface = "bold", 
            check_overlap = FALSE) + 
  annotate(geom = "text", 
           x = -90, 
           y = 26, 
           label = "Gulf of Mexico", 
           fontface = "italic", 
           color = "grey22", 
           size = 6) +
  annotation_scale(location = "bl", 
                   width_hint = 0.5) + 
  annotation_north_arrow(location = "bl", 
                         which_north = "true", 
                         pad_x = unit(0.75, "in"), 
                         pad_y = unit(0.5, "in"), 
                         style = north_arrow_fancy_orienteering) +
  coord_sf(xlim = c(-102.15, -74.12), 
           ylim = c(7.65, 33.97), 
           expand = FALSE) + 
  xlab("Longitude") + 
  ylab("Latitude") + 
  ggtitle("Map of the Gulf of Mexico and the Caribbean Sea") + 
  theme(panel.grid.major = element_line(color = gray(.5), 
                                        linetype = "dashed", 
                                        size = 0.5), panel.background = element_rect(fill = "aliceblue")
  )
```

 