---
title: "Lab 3 notes: Data tidying with `tidyr`"
toc: true
format:
  html:
    code-overflow: wrap
  pdf:
    keep-tex: true
    include-in-header: 
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
        \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
execute:
  class-output: OutputCode
embed-resources: true
editor_options: 
  chunk_output_type: console
---

<!-- This document guides you through the lab today. For the lab assignment, you'll make a separate Quarto document, type up and solve some exercises mentioned throughout the lab document here, and render and submit that document (not this lab notes document). -->

This document guides you through the lab today. As you work through it, this document will point out a good time to work on particular exercises in the separate Lab 3 assignment document, `lab03_tidyr.qmd`. For your submission on Gradescope, render and submit the Lab 3 assignment document, not this lab notes document.

## First, from last time: summary statistics (Sec 3.5)

Tidy datasets have one row per observation and one column per variable. Recall that in the flights dataset, each row (observation) represents a flight, and each column (variable) represents some piece of information about that flight: departure delay, month, scheduled departure time, and so on.

Say that you want to know the average delay by month in the `flights` dataset.  Or say that you want to know the shortest and longest delay for flights to each destination. Both of these are summary statistics that require us to

1. group the dataset by a variable (month for the first problem, destination for the second), then
2. compute some summary statistic (minimum, maximum, mean, sum, etc.) of the variable(s) of interest for all observations (which are flights here) in each group.

This means that the result we get should have one row per *group* (like a month), not individual observation (like a flight), and each row should tell us what group that row represents and what the summary statistics are for that group.

Here's those two examples, written out in code. FIRST, what code do you have to run before the code below will run? Add that code to a new code chunk at the top of this lab for good practice.

```{r}
#| eval: false

flights |> # start with the dataset
  group_by(month) |> # group by month so that you'll compute whatever it is you want for each month
  summarize( # here's where we compute however many summary statistics we want
    avg_delay = mean(dep_delay) # this is the average departure delay
  )

flights |> 
  group_by(dest) |> 
  summarize( # here we compute more than one summary statistic
    shortest_delay = min(dep_delay),
    longest_delay = max(dep_delay)
  )
```

Notice that a lot of the output just says `NA`, which means the value is missing. This is because we have included missing values in our calculations. For example, try running the following code:

```{r}
mean(c(1, 3, NA)) # returns NA because there is a missing value in the data
mean(c(1, 3, NA), na.rm = TRUE) # takes the mean of only non-missing values, which here are 1 and 3
```

Many functions other than `mean` have the same `na.rm` argument that allows you to remove NAs (hence it's called `na.rm` for "NA remove"). The default is FALSE.

So let's retry the code above with this new argument:

```{r}
#| eval: false

flights |> # start with the dataset
  group_by(month) |> # group by month so that you'll compute whatever it is you want for each month
  summarize( # here's where we compute however many summary statistics we want
    avg_delay = mean(dep_delay, na.rm = TRUE) # this is the average departure delay
  )

flights |> 
  group_by(dest) |> 
  summarize( # here we compute more than one summary statistic
    shortest_delay = min(dep_delay, na.rm = TRUE),
    longest_delay = max(dep_delay, na.rm = TRUE)
  )
```

Before we move on, make some notes to yourself (not graded, just for you) about what patterns you notice here.

1. How can you tell what variable to group by?

2. How many rows and columns are there in the output of `summarize` in each of these examples?

3. In these examples, which functions is na.rm an argument of? In other words, where did you put it when we needed it: in the `group_by` function, the `summarize` function, the `min` function, the `max` function, the `mean` function...?

4. What part of the code specifies the name of the summary statistic you're computing (not the function you use to compute it, but the name of the column in the output that shows the results of what you computed)? For instance, in the second example, where did you name the column of the output that will have the minimum departure delay?

Then try your hand at **Exercises 1-2 in the Lab 3 assignment**.

Later if you're interested and you haven't already done so, you can check out Section 3.5 to learn more about grouping datasets and to see another way to use `summarize`.



## Tidying data

Today we're going to work through the content of Chapter 5 of R4DS2e, but without following the textbook exactly.

Take a look at these two datasets (see `?table1` to learn more about what these data represent):

```{r}
#| eval: false
table1
table2
```

In Tables 1-2, we're looking at the same data in a different layout. We always have the same three countries, two years, and two variables. In each table, how many rows and columns do you have in each table? How does this relate to the number of countries, years, and variables?

Of these two tables, Table 1 is the more natural form for this data if we're interested in a table. It's also great if we want to plot a (very short) time series for each country for a single variable, say, `cases`. **Make this plot in Exercise 3 of `lab03_tidyr.qmd`.**

What if we want to plot both variables? For instance, we might want to make side-by-side plots of `cases` and `population`. To do this, we can use table2:

```{r}
#| eval: false
ggplot(table2, aes(x = year, y = count, group = country, color = country)) +
  geom_line() +
  facet_wrap(~ type, scales = "free")
```

Basically, having one row per value and having a column that tells us what variable that value represents is a format that works well for many more complex, multi-variable plots.

So what if we're given `table1` but we want to make `table2` for use in a plot? Enter pivoting!

There are two basic ways to pivot: we either want to `pivot_longer` (end up with more rows, typically fewer columns) or `pivot_wider` (end up with fewer rows, typically more columns). Here, how do we know which one we want in order to go from `table1` to `table2`?

```{r}
#| eval: false
mytable2 = pivot_longer(
  table1,
  cols = cases:population,
  names_to = "type",
  values_to = "count"
  )
```

Carefully compare `mytable2` to `table2` to check if they are the same.

What if you want to go backwards, meaning you have `table2` and you want to get `table1`?

```{r}
#| eval: false
mytable1 = pivot_wider(
  table2,
  names_from = "type",
  values_from = "count"
  )
```

Work on **Exercises 4-5 in the Lab 3 assignment**

[This pivoting vignette in the R documentation](https://tidyr.tidyverse.org/articles/pivot.html) is the best compilation I've seen of varied examples that show a wide breadth of different cases in which you can use pivoting. Go through each of these examples, reading the text and trying the code yourself.

Then **finish the rest of the Lab 3 assignment.**



