---
title: "STAT 408 Lab 1: Data Viz with `ggplot`"
author: "Hadley Stabnau"
date: today
date-format: "January 15, 2026"
format: pdf
embed-resources: true
---

1. Load the three packages you are using for this lab on the `penguins` dataset.

```{r}
library(tidyverse)
library(palmerpenguins)
library(ggthemes)
```



2. How many rows (observations) are in `penguins`? How many columns (variables)? What are the variables (column names) in this dataset? Answer these three questions using code covered in **lab01_notes.qmd**.

```{r}
nrow(penguins)
ncol(penguins)
```
There are 344 observations (rows) in the penguin data set and 8 variables (columns).


3. What does the `bill_depth_mm` variable in the `penguins` data frame describe? Read the help for ?penguins to find out.

The bill_depth_mm variable in this data frame describes the bill depth in mm of a given penguin.



4. Make a scatterplot of `bill_depth_mm` vs. `bill_length_mm.` That is, make a scatterplot with `bill_depth_mm` on the y-axis and `bill_length_mm` on the x-axis. Describe (in text, not code or code comments) the relationship between these two variables.

```{r}
# Code chunk 1
ggplot(
  data = penguins,
  mapping = aes(x = bill_depth_mm, y = bill_length_mm)
) +
  geom_point()
```

These two variables, bill depth and bill length, appear to have a weak positive linear relationship with two possible different groups. There do not appear to be any outliers in this data plot.


5. What happens if you make a scatterplot of `species` vs. `bill_depth_mm`? (Make and show this scatterplot, then discuss in text what happens.) What might be a better choice of geom for this plot? (See other options mentioned in Section 1.2.3.)

```{r}
# Code chunk 1
ggplot(
  data = penguins,
  mapping = aes(x = species, y = bill_depth_mm)
) +
  geom_point()
```

A better choice of geom for this plotted data would be some kind of bar graph. This is because each species will lie in the same bar/line anyways, and a bar graph would show the differences in bill depths by the height of each bar. 



6. Why does the following code give an error? (Write your answer in text.) Fix it. (NOTE: you need to delete the line that says `#| eval: false` so that this code will run when you render your Quarto document.)

The error in this chunk of code is that there is no variable assigned to the x or the y axis', and there is not any specific mapping assigned so R has the correct data but doesn't know what we want it to do with that data. To fix this, I just stole the variables that were used in the previous question.

```{r}


ggplot(data = penguins,mapping = aes(x = species, y = bill_depth_mm)) + 
  geom_point()
```



7. What does the `na.rm` argument do in `geom_point()`? What is the default value of the argument? (Hint: to answer these questions, look up the help page for `geom_point()` the same way you did for the `penguins` dataset, and don't include the parentheses when you do that.) Then create a scatterplot where you successfully use this argument set to `TRUE`.

The na.rm arument in geom_point() removes missing values, whether or not we set na.rm equal to true or false determines if the values are removed silently or with a warming message

```{r}

ggplot(
  data = penguins,
  mapping = aes(x = bill_depth_mm, y = bill_length_mm)
) +
  geom_point(na.rm=FALSE)
```



8. Recreate the visualization in Exercise 8 of Section 1.2.5. What aesthetic should `bill_depth_mm` be mapped to? And should it be mapped at the global level or at the geom level?

Bill depth will be mapped to the color aesthetic. This is at the geom level not the global level.

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(mapping = aes(color = bill_depth_mm)) +
  geom_smooth(method = "gam")
```



9. Add the following caption to the plot you made in the previous exercise: "Data come from the palmerpenguins package." Hint: Take a look at the documentation for` labs()`.

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(mapping = aes(color = bill_depth_mm)) +
  geom_smooth(method = "gam") +
  labs(caption="Data come from the palmerpenguins package")
```



10. Will these two graphs look different? Why/why not? (Please leave the `#| eval: false` option in this code chunk.)

```{r}
#| eval: false

ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point() +
  geom_smooth()

ggplot() +
  geom_point(
    data = penguins,
    mapping = aes(x = flipper_length_mm, y = body_mass_g)
  ) +
  geom_smooth(
    data = penguins,
    mapping = aes(x = flipper_length_mm, y = body_mass_g)
  )
```

These two graphs will look the same because even though their code is not identical, in the first plot the data is set globally which automatically sort of carries down to the geom point/smooth. Whereas in the second plot, the data is directly set inside of the geom point and geom smooth. So overall, both of those code chunks are building the same plots, but just applying the data in different ways within the code. 



11. Type the name of the function we used in this lab for each of the following tasks.

- Getting the number of variables in a dataset: *ncol()*
- Setting up plot aesthetics: *mapping=aes()*
- Loading a package: *library()*
- Getting the number of rows in a dataset: *nrow()*
- Installing a package: *install.packages()*



12. Make a few notes here for yourself about some useful tips and tricks you learned (or were reminded of) in today's lab.

The tip I learned/reminded myself of today that was incredibly helpful was the read in help function, being able to just pull up information on my screen on pretty much anything in R is awesome. The Label function is also a new function I learned today that will certainly come in handy.




