---
title: "STAT 408 Lab 2: dplyr"
author: "Hadley Stabnau"
date: "Jan 22, 2026"
format: pdf
---

## Section 3.2.5

1. In a single pipeline for each condition, find all the flights that meet the condition:

* Had an arrival delay of two or more hours: 10,200 flights
* Flew to Houston (IAH or HOU): 9,313 flights
* Were operated by United, American or Delta: 139,504 flights
* Departed in summer (July, August, and September): 86,995 flights
* Arrived more than two hours late but didn’t leave late: 210,094 flights
* Were delayed by at least an hour, but made up over 30 minutes in flight: 45,380 flights

```{r, message=FALSE}
#install.packages("nycflights13")
library(nycflights13)
library(tidyverse)
glimpse(flights)

flights |>
  count(arr_delay>=120)

flights |> 
  filter(dest=="IAH" |dest=="HOU") |>
  count(dest)

flights |>
  filter(carrier=="UA" |carrier=="AA" |carrier=="DL") |>
  count(carrier)

flights |>
  filter(month==6 |month==7 |month==8) |>
  count(month)

flights |>
  count(arr_delay>120 |dep_delay<=0)

flights |>
  count(dep_delay>=60 |(dep_delay-arr_delay)>=30)


```

2. Sort `flights` to find the flights with the longest departure delays. Find the flights that left earliest in the morning.

```{r}
flights |>
  arrange(desc(dep_delay))

flights |>
  arrange(dep_time)
```



3. Sort `flights` to find the fastest flights. (Hint: Try including a math calculation inside of your function).

```{r}


flights|>
  arrange(arr_time-dep_time)
```

4. Was there a flight every day of 2013?

```{r}
flights |>
  distinct(year, month, day) |>
  count()

```
Yes, according to the above table, there are 365 days out of the year 2013 that a flight occured. 


5. Which flights traveled the farthest distance? Which traveled the least distance?

```{r}
flights |>
  arrange(desc(distance))

flights |>
  arrange(distance)


```

6. Does it matter what order you used the `filter()` and `arrange()` if you're using both? Why/Why not? Think about the results. and how much work the functions would have to do. 

It does not matter what order you use `filter()` and `arrange()` when using both because either way, the data will be filtered out for only the observations we're interested in and arranged in some order. Whether you filter the data out first or not, shouldn't matter. See the example below where we filter the data so that we are only looking at flights from February and we are looking at the flights with the shortest air times:
```{r}
flights |>
  filter(month==2) |>
  arrange(air_time)

flights |>
  arrange(air_time) |>
  filter(month==2)
```


## Section 3.3.5

1. Compare `dep_time`, `sched_dep_time`, and `dep_delay`. How would you expect those three numbers to be related?

I would expect these. three numbers to be related because when `dep_delay` increases, that increase will be equal to the difference between `sched_dep_time` and `dep_time`.

```{r}
flights |>
  arrange(flight)
```

2. Brainstorm as many ways as possible to select `dep_time`, `dep_delay`, `arr_time`, and `arr_delay` from flights.

`select()` and `relocate()` are different ways I've thus far learned to pick up or change the position of the columns. An example of using them would be:
```{r}
flights |>
  select(dep_time, dep_delay, arr_time, arr_delay)

flights|>
  relocate(dep_time, dep_delay, arr_time, arr_delay)

```

3. What happens. if you specify the name of the same variable multiple times in a `select()` call?

```{r}
flights |>
  select(dep_time, arr_time, dep_time)
```

If you specify the same variable name multiple times in a select call, R just ignores the repeat and continues with the rest as usual. 



4. What does the any_of() function do? Why might it be helpful in conjunction with this vector?

```{r}
variables <- c("year", "month", "day", "dep_delay", "arr_delay")
```

The any_of() function is a selection helper which does not check for missing variables, but it does help with negative selections (making sure a variable is removed). This may be helpful for conjunction with the above vector, because it can remove or ignore the variables being asked for that aren't in the given data set. 



5. Does the result of running the following code surprise you? How do the select helpers deal with upper and lower case by default? How can you change that default?

```{r}
flights |> select(contains("TIME"))
```

The result of running that code did not surprise me, it appears to have grabbed/selected the columns that contain the word "time" in them. The select helpers are case sensitive. To change this, you can use certain commands like `ignore_case=TRUE` which will ignore/match the characters to upper or lower case, whichever is neccessarey. Ignore_case came from section 15.5 of the textbook. 



6. Rename air_time to air_time_min to indicate units of measurement and move it to the beginning of the data frame.

```{r}
flights |>
  rename(air_time_min = air_time) |>
  relocate(air_time_min)


```


7. Why doesn't the following work, and what does the error mean?

```{r}
#flights |> 
  #select(tailnum) |> 
  #arrange(arr_delay)
#> Error in `arrange()`:
#> ℹ In argument: `..1 = arr_delay`.
#> Caused by error:
#> ! object 'arr_delay' not found
```


This code doesn't work because `select` drops all of the other columns except `tailnum` so then when we go to `arrange(arr_delay)`, it no longer exists because we have already filtered out all of the columns except `tailnum` and there is no longer a variable to pick out called `arr_delay` in this code chunk.

I have commented out that last code chunk because R will not render my document with it. 

