---
title: "STAT 408 Lab 07: Logicals"
author: "Hadley Stabnau"
date: "Feb 26, 2026"
format: html
editor: visual
---

## Read in Data


```{r, message=FALSE, echo=TRUE}
library(tidyverse)

Bozeman <- read.csv("Living in Bozeman.csv")
names(Bozeman)



```

```{r, echo=TRUE, message=FALSE}
names(Bozeman) = c("timestamp","years_live_boz", "for_fun_boz", "part_boz_live", "fav_thing_boz_why", "fav_season_boz", "fav_least_rest_boz", "home_to_campus_miles_boz", "msu_credits")


```

## Explore the Data

* The variables with a limited set of fixed possible areas are: `years_live_boz`, `fav_season_boz`. This is because these are both the multiple choice questions. 
```{r}
table(Bozeman$years_live_boz)
prop.table(table(Bozeman$years_live_boz))


Bozeman |>
  ggplot(aes(x=years_live_boz)) +
  geom_bar() +
  labs(title = "Years Lived in Bozeman", x="Years")
```
```{r}
table(Bozeman$fav_season_boz)
prop.table(table(Bozeman$fav_season_boz))


Bozeman |>
  ggplot(aes(x=fav_season_boz)) +
  geom_bar() +
  labs(title = "Favorite Season in Bozeman", x= "Seasons")
```
* Numeric variables are `home_to_campus_miles_boz` and `msu_credits`.
```{r}
colnames(Bozeman)
```

```{r, warning=FALSE}
Bozeman$home_to_campus_miles_boz <-
  as.numeric(gsub("[^0-9.]", "", Bozeman$home_to_campus_miles_boz))

Bozeman$msu_credits <-
  as.numeric(gsub("[^0-9.]", "", Bozeman$msu_credits))

Bozeman |>
  ggplot(aes(x= home_to_campus_miles_boz)) +
  geom_density() +
  labs(title = "Distance from Home to Campus")


Bozeman |>
  ggplot(aes(x= msu_credits)) +
  geom_density() +
  labs(title = "Credits Currently Taken at MSU")


```
* The plot showing the distribution of the distance in miles from ones home to campus has a fairly normal shaped but skewed distribution, it all has only one mode, it doesn't appear to have any extreme outliers, and data is spread out around an mean of 1 mile.

* The plot showing the distribution of ones msu credits is severely right skewed meaning there may be one or two extreme values in the data causing it to be pulled to the right. There does appear to be multiple modes and the data peaks around a mean of 15 credits.


* The columns with missing data are `fav_least_rest_boz` and `part_boz_live`. The pattern I am seeing in the NA or missing values, is that those taking the questionaire either answered one or the other of the questions but not both and not neither. 

* It's possible that there are outliers as both of the distributions I plotted are skewed but without doing th IQR test, we can't know for sure. 

## Manipulate the Data


```{r}
Bozeman$credits <- ifelse(Bozeman$msu_credits > 15, "TRUE", "FALSE")
```
* The above uses the `ifelse()` function to display a new column called `credits` and this displays a True if the column meets the requirement of >15 credits and False if it does not.

```{r}
Bozeman <- Bozeman |>
  mutate(for_fun_bozo = case_when(
    for_fun_boz %in% c("Hiking", "hike", "Hike") ~ "Hike"
  )) 
```

```{r, message=FALSE}
library(tidyverse)

Bozeman <- Bozeman |>
  mutate(distance_travel_campus = case_when(
   home_to_campus_miles_boz < 1 ~ "Close",
   home_to_campus_miles_boz >=1 & home_to_campus_miles_boz < 5 ~ "Medium",
   home_to_campus_miles_boz >= 5 ~ "Far"
  ))

Bozeman
```

