---
title: "STAT 408 Lab 08: Strings and Text"
author: "Hadley Stabnau"
date: "Mar 08, 2026"
format: html
editor: visual
---

```{r}
options(tibble.print_max = 100)
library(tidyverse)
library(dplyr)
```

## Problem 1

```{r, message=FALSE, echo=FALSE}
stat408 <- read_csv("/Users/hadleystabnau/Desktop/STAT408/data/408 Survey.csv")
names(stat408)

names(stat408) = c("timestamp", "dist_to_campus", "where_from", "fav_class", "edrinks_per_wk", 
                   "fav_dining", "fav_NFL", "major", "year", "study_hours_per_wk",
                   "fav_day", "fav_season", "fav_water_pH", "time_to_campus", "hrs_sleep_per_day",
                   "fav_olympic")
names(stat408)
```

```{r, message=TRUE, echo=TRUE}
stat408 |>
  separate(fav_olympic,
           into = c("favorite sport", "reason"),
           sep = " because ", 
        )
```

There are still a few rows that have NA in the reason column and that is because those are the rows that didn't specifically say "because" when justifying their favorite olympic sport. However, the command separated the row called "fav_olympic" into two, one called "fav_sport" and one called "reason" separated in each row by the appearance of the word "because".

\newpage

## Problem 2

```{r, message=FALSE, echo=FALSE}
sports <- read_csv("/Users/hadleystabnau/Desktop/STAT408/data/STAT 408 Lab 7 Sports Survey.csv")

names(sports)

names(sports) = c("timestamp", "fav_sport", "goat", "num_sports_played", "fav_jersey_num", 
                   "fav_sports_team", "amount_sports_affect_life", "best_sports_drink", "avg_miles_run_soccer_game")

names(sports)
```

```{r, echo=FALSE}
sports <- sports |>
  mutate(best_sports_drink = case_when(
    str_detect(best_sports_drink, regex("gatorade", ignore_case = TRUE)) ~ "Gatorade",
    TRUE ~ best_sports_drink
  ))


sports <- sports |>
  mutate(best_sports_drink = case_when(
    str_detect(best_sports_drink, regex("body armor", ignore_case = TRUE)) ~ "Body Armor",
    TRUE ~ best_sports_drink
  ))

sports <- sports |>
  mutate(best_sports_drink = case_when(
    str_detect(best_sports_drink, regex("powerade", ignore_case = TRUE)) ~ "Powerade",
    TRUE ~ best_sports_drink
  ))

sports <- sports |>
  mutate(best_sports_drink = case_when(
    str_detect(best_sports_drink, regex("water", ignore_case = TRUE)) ~ "Water",
    TRUE ~ best_sports_drink
  ))


 sports <- sports |>
  mutate(best_sports_drink = case_when(
    str_detect(best_sports_drink, regex("Gin and Tonic", ignore_case = TRUE)) ~ "Other",
    TRUE ~ best_sports_drink
  ))
 
 sports <- sports |>
  mutate(best_sports_drink = case_when(
    str_detect(best_sports_drink, regex("Grain alcohol", ignore_case = TRUE)) ~ "Other",
    TRUE ~ best_sports_drink
  ))
 

```

\newpage

## Problem 3

```{r}
school <- read_csv("/Users/hadleystabnau/Desktop/STAT408/data/School_survey.csv")

names(school) = c("timestamp", "major", "number_credits", "houirs_hw_per_week", "fav_class", "num_semesters_grad", "cats_or_griz", "extracurriculars", "extracurriculars_pos_neg", "extracurriculars_before_college", "met_friend", "met_enemy", "live_together", "effect_on_schoolwork", "hours_week_extracurricular", "money_spent_extracurricular")


```

```{r, echo=TRUE, message=TRUE}
school <- school |>
  separate_rows(extracurriculars, sep = "&") |>
  separate_rows(extracurriculars, sep = ",")
```

\newpage

## Problem 4

```{r}
bozeman <- read_csv("/Users/hadleystabnau/Desktop/STAT408/data/Living in Bozeman.csv") 

names(bozeman) = c("timestamp","years_live_boz", "for_fun_boz", "part_boz_live", "fav_thing_boz_why", "fav_season_boz", "fav_least_rest_boz", "home_to_campus_miles_boz", "msu_credits")
```

```{r, echo=TRUE, message=FALSE}
bozeman <- bozeman |>
  mutate(part_boz_live = case_when(
    str_detect(part_boz_live, regex("campus", ignore_case = TRUE)) ~ "Near Campus",
    TRUE ~ part_boz_live
  ))

bozeman <- bozeman |>
  mutate(part_boz_live = case_when(
    str_detect(part_boz_live, regex("north", ignore_case = TRUE)) ~ "North",
    TRUE ~ part_boz_live
  ))

bozeman <- bozeman |>
  mutate(part_boz_live = case_when(
    str_detect(part_boz_live, regex("south", ignore_case = TRUE)) ~ "South",
    TRUE ~ part_boz_live
  ))

bozeman <- bozeman |>
  mutate(part_boz_live = case_when(
    str_detect(part_boz_live, regex("babcock", ignore_case = TRUE)) ~ "Central",
    TRUE ~ part_boz_live
  ))

bozeman <- bozeman |>
  mutate(part_boz_live = case_when(
    str_detect(part_boz_live, regex("19th", ignore_case = TRUE)) ~ "Central",
    TRUE ~ part_boz_live
  ))
```

\newpage

## Problem 5

```{r}
bozeman |>
  separate(fav_thing_boz_why,
           into = c("favorite thing about bozeman", "reason"),
           sep = " because ", 
        )
```

I wanted to separate th column askingn what is your favorite thing about Bozeman and why? into two different columns of the thing and the reason. The code I used did work for some of the rows of this variable, but due to the wording of the responses, there are stil a lot of rows that need help separating.

\newpage

## Problem 6

Two text variables with cleaning problems that I do not know how to correct are; having numerical values get rid of labels like "miles" or "lbs", and having r fill a place with a specific word when there is a missing value. To avoid this issue in the first place, they can be more specific in the study design as to whether or not they want respondents to include labels on numerical responses. Also, providing some sort of word bank for questions or a guide for responses that we want to look a certain way.
